**Title:** Analyzing long-term water quality of lakes in Rhode Island and the Northeastern United States with an anomaly approach

### 2021-03-19 manuscript/ecosphere_production/manuscript.Rmd
finally final?!?

URIWW data are included in the LAGOSNE and contributed to the larger regional study by Oliver et al. [-@Oliver_2017], which also allowed us to examine the long-term trends specifically in Rhode Island.  The URIWW program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring locations on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers. URIWW now provides more than 90% of Rhode Island's lake baseline data and is an integral part of the state's environmental data collection strategy. Data quality assurance and control is treated with paramount importance; volunteers are trained both in the classroom and the field, regular quality checks occur, and volunteers are provided with all the necessary equipment and supplies, along with scheduled collection dates. For freshwater lakes and reservoirs, weekly Secchi depth and water temperature are recorded, along with bi-weekly chlorophyll *a* and in deep lakes (greater than 5 meters) dissolved oxygen. Water samples are collected three times per season (May through October) to be analyzed for nutrients and bacteria.

@@

Prior studies have modeled water quality trends across a larger region of the northeastern US within 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017]. We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by Oliver et al. [-@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne; @sorannolagosne2017; @sorannolimno2019]. Temperature data were not available, thus we examined trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.  We used the same selection criteria on the LAGOSNE dataset as was applied to the URIWW data. This ensured that both datasets represented the same seasonal time frame.  Exact measurement depth for the LAGOSNE data were not available; however, the water quality data are reported to be either surface or epilimnion samples [@sorannolagosne2017]. At the time of our analysis, the LAGOSNE data do not extend beyond 2013 thus sites needed to have a sampling event in 1993-2002 and in 2003-2013 to be included.

@@

Lastly, data and code for the entire analysis is available from <https://github.com/USEPA/ri_wq_trends/> and is archived at <https://doi.org/10.5281/zenodo.4306765> [@zenodo_citation].  All analyses were conducted with R version 4.0.3 and details on R package versions and operating system used for this analysis are included in a file, `sessioninfo.txt` at <https://github.com/USEPA/ri_wq_trends/blob/master/sessioninfo.txt> [@r2021; @lubridate; @broom; @stringr; @dplyr; @readr; @readxl; @tidyr; @here; @ggplot2; @hrbrthemes; @kendall; @sf; @usaboundaries; @usaboundariesdata; @knitr; @captioner; @pander; @cowplot; @tidycensus; @feddata; @nsink; @units; @xie2015dynamicdoc; @magick; @flextable].  Values included in each of the figures have also been separately saved as a comma-separated values file, `yearly_average_anomaly.csv`, and may be retrieved via <https://github.com/USEPA/ri_wq_trends/blob/master/data/yearly_average_anomaly.csv>.

@@

There are many different methods for analyzing time series data for trends.  Environmental data are notoriously "noisy" and one of the difficulties that is encountered with multiple sampling sites is how to identify a trend while there is variation within a sampling location as well as variation introduced by differing start years for sampling among the many sites.  For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies, then a simple comparison of raw values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more degraded water bodies.  Thus, it is necessary to account for this type of within-site and among-site variation, using methods similar to those used to analyze long-term temperature trends using temperature anomalies [e.g., @jones1996calculating].  The general approach we used calculates site-specific deviations from a long-term median over a pre-determined reference period.  This is slightly different than the typical use with temperature anomalies as those usually use the mean instead of the median.  Many of the variables we looked at are non-normal and often have outliers (e.g. algal blooms).  In this case the median is preferred. Using the median with site-specific deviations allows all sites to be shifted to a common baseline and the deviations, or anomalies, indicate change over the specified reference period.  We refer to this method as "site-specific anomalies".

@@

Our sub-regional analysis indicates that increases in primary production, as measured by chlorophyll *a*, occurred over the ~25 years of our study period.  Over the same period we also demonstrate long-term warming of Rhode Island lakes and reservoirs. Chlorophyll has increased, on average, `r signif(ww_chla_rl$slope, 2)` μg/L per year over the 23 years of our analysis, while temperature has increased `r signif(ww_temp_rl$slope, 2)` °C per year over the same period.  Trends in nutrients suggested little change. Total nitrogen, total phosphorus, and the nitrogen:phosphorus ratio all showed no evidence of trends.   While our analysis is not capable of detecting causation, that both chlorophyll *a* and temperature is increasing with less obvious trends in nutrients is interesting and warrants further exploration to see if increasing chlorophyll *a* can be described by temperature.  Also, geographic extent does indeed matter when trying to identify long-term water quality trends. Similar to the results of Oliver et al. [-@Oliver_2017] our analysis shows little increasing trend in chlorophyll *a* at the regional scale (e.g., northeastern and mid-western United States).  However, at the local scale of the state of Rhode Island, there is a clear increasing trend in chlorophyll *a*.

@@

There are several broader implications from the results of our analysis and of examining long-term water quality trends in general. First, as more long-term datasets become available, it is important for managers, stakeholders, and researchers to work together to better understand long-term water quality trends at multiple spatial scales.  Specifically for this study, the results provide feedback to long time volunteer monitors about the trends in long-term, volunteer collected data.  This type of feedback is important in maintaining involvement as it has been shown in other areas that contributing to science and management is one of the expectations for some community science volunteers and can help maintain longer-term involvement [@ganzevoort2017sharing; @bonney2009citizen; @gouveia2004promoting].  If long-term involvement was not common, then consistent long-term data may be more difficult to obtain and understanding of long-term trends would be made more difficult.

>>>>

The URIWW program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring locations on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers. URIWW now provides more than 90% of Rhode Island's lake baseline data and is an integral part of the state's environmental data collection strategy. Data quality assurance and control is treated with paramount importance; volunteers are trained both in the classroom and the field, regular quality checks occur, and volunteers are provided with all the necessary equipment and supplies, along with scheduled collection dates. For freshwater lakes and reservoirs, weekly Secchi depth and water temperature are recorded, along with bi-weekly chlorophyll *a* and in deep lakes (greater than 5 meters) dissolved oxygen. Water samples are collected three times per season (May through October) to be analyzed for nutrients and bacteria.

@@

Prior studies have modeled water quality trends across a larger region of the northeastern US within 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island (which includes the URIWW data), Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017]. We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by Oliver et al. [-@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne; @sorannolagosne2017; @sorannolimno2019]. Temperature data were not available, thus we examined trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.  We used the same selection criteria on the LAGOSNE dataset as was applied to the URIWW data. This ensured that both datasets represented the same seasonal time frame.  Exact measurement depth for the LAGOSNE data were not available; however, the water quality data are reported to be either surface or epilimnion samples [@sorannolagosne2017]. At the time of our analysis, the LAGOSNE data do not extend beyond 2013 thus sites needed to have a sampling event in 1993-2002 and in 2003-2013 to be included.

@@

Lastly, data and code for the entire analysis is available from <https://github.com/USEPA/ri_wq_trends/> and is archived at <https://doi.org/10.5281/zenodo.4306765> [@zenodo_citation].  All analyses were conducted with R version 4.0.4 and details on R package versions and operating system used for this analysis are included in a file, `sessioninfo.txt` at <https://github.com/USEPA/ri_wq_trends/blob/master/sessioninfo.txt> [@r2021; @lubridate; @broom; @stringr; @dplyr; @readr; @readxl; @tidyr; @here; @ggplot2; @hrbrthemes; @kendall; @sf; @usaboundaries; @usaboundariesdata; @knitr; @captioner; @pander; @cowplot; @tidycensus; @feddata; @nsink; @units; @xie2015dynamicdoc; @magick; @flextable].  Values included in each of the figures have also been separately saved as a comma-separated values file, `yearly_average_anomaly.csv`, and may be retrieved via <https://github.com/USEPA/ri_wq_trends/blob/master/data/yearly_average_anomaly.csv>.

@@

There are many different methods for analyzing time series data for trends.  Environmental data are notoriously "noisy" and one of the difficulties that is encountered with multiple sampling sites is how to identify a trend while there is variation within a sampling location as well as variation introduced by differing start years for sampling among the many sites.  For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies, then a simple comparison of raw values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more degraded water bodies.  Thus, it is necessary to account for this type of within-site and among-site variation.  To do so, we used methods similar to those used to analyze long-term temperature trends using temperature anomalies [e.g., @jones1996calculating].  The general approach we used calculates site-specific deviations from a long-term median over a pre-determined reference period.  This is slightly different than the typical use with temperature anomalies as those usually use the mean instead of the median.  Many of the variables we looked at are non-normal and often have outliers (e.g. algal blooms).  In this case the median is preferred. Using the median with site-specific deviations allows all sites to be shifted to a common baseline and the deviations, or anomalies, indicate change over the specified reference period.  We refer to this method as "site-specific anomalies".

@@

Our sub-regional analysis indicates that increases in primary production, as measured by chlorophyll *a*, occurred over the ~25 years of our study period.  Over the same period we also demonstrate long-term warming of Rhode Island lakes and reservoirs. Chlorophyll has increased, on average, `r signif(ww_chla_rl$slope, 2)` μg/L per year over the 23 years of our analysis, while temperature has increased `r signif(ww_temp_rl$slope, 2)` °C per year over the same period.  Trends in nutrients suggested little change. Total nitrogen, total phosphorus, and the nitrogen:phosphorus ratio all showed no evidence of trends.   While our analysis is not capable of detecting causation, that both chlorophyll *a* and temperature is increasing with less obvious trends in nutrients is interesting and warrants further exploration to see if increasing chlorophyll *a* can be described by temperature.  Also, geographic extent does indeed matter when trying to identify long-term water quality trends. Similar to the results of Oliver et al. [-@Oliver_2017] our analysis shows little increasing trend in chlorophyll *a* at the regional scale (e.g., northeastern and mid-western United States).  However, at the more localized scale of the state of Rhode Island, there is a clear increasing trend in chlorophyll *a*.

@@

There are several broader implications from the results of our analysis and of examining long-term water quality trends in general. First, as more long-term datasets become available, it is important for managers, stakeholders, and researchers to work together to better understand long-term water quality trends at multiple spatial scales.  Specifically for this study, the results provide feedback to long time volunteer monitors about the trends in long-term, volunteer collected data.  This type of feedback is important in maintaining involvement as it has been shown in other areas that contributing to science and management is one of the expectations for some community science volunteers and can help maintain longer-term involvement [@ganzevoort2017sharing; @bonney2009citizen; @gouveia2004promoting].  If long-term involvement was not common, then consistent long-term data may be difficult to obtain and understanding of long-term trends would be challenging.
